### 2017-03-24 concl.tex
exp, conc

\section{Conclusion}
The approximate linear programming (ALP) is an approximate dynamic programming method that addresses the prediction and control problems successfully. However, an important shortcoming of the ALP is that it has large number of constraints, which is tackled in practice by sampling a tractable number of constraints from the ALP to formulate and solve a reduced linear program (RLP). Though RLP has been found to work well empirically in various domains ranging from queues to Tetris games, performance guarantees are available only in the case of a specific RLP formulated under idealized assumptions. Thus there has been a gap in the theory of constraint reduction.\\
In this paper, we introduced a novel framework based on the generalized reduced linear program formulation to study constraint reduction. The constraints of the GRLP were obtained as positive linear combinations of the original ALP. We provided an error bound that relates the optimal value function to the solution of the GRLP. Our error bound contained two terms, one inherent to the ALP formulation and the other due to constraint reduction. We also made qualitative and quantitative observations about the nature of the error term that arose due to constraint reduction. Our analysis also revealed the fact that it is not always necessary to sample according to the stationary distribution of the optimal policy and, in fact, potentially several different constraint sampling/approximation strategies might work. In particular, we also theoretically justified linear function approximation of the constraints. We also discussed the results and provided a numerical example in the domain of controlled queues. To conclude, we observe that by providing a novel theoretical framework to study constraint approximation, this paper provides important results that add to the theory of ALP.

>>>>

%!TEX root =  autocontgrlp.tex
\section{Conclusion}
In this paper, we introduced and analyzed the linearly relaxed approximate linear program (LRALP) whose constraints were obtained as positive linear combination of the original constraints of the ALP. 
The main novel contribution is a theoretical result which gives a geometrically interpretable bound on the performance loss due to relaxing the constraint sets. Possibilities for future work include extending the results to other forms of approximate linear programming in MDPs (e.g., \citep{SALP}), designing algorithms that use the newly derived results to actively compute what constraints to select, just to mention a few.
%By providing performance analysis for the GRLP, this paper justified linear function approximation of the dual variables of the ALP. 
\begin{comment}
The approximate linear program (ALP) is a widely employed method to handle MDPs with large number of states. However, an important shortcoming of the ALP is that it has large number of constraints, which is tackled in practice by sampling a tractable number of constraints from the ALP to formulate and solve a reduced linear program (RLP). The RLP has been found to work well empirically in various domains ranging from queues to Tetris games, and performance guarantees are for a specific RLP formulated under idealized assumptions. Alternatively, function approximation in the dual variables of the ALP has been another approach that has been employed in literature \cite{ALP-Bor,dolgov} to address the issue of large number of constraints. However there was no prior theoretical characterization for the function approximation of the dual variables.\par
In this paper, we introduced a novel framework based on the generalized reduced linear program formulation. The constraints of the GRLP were obtained as positive linear combinations of the original ALP. We provided an error bound that relates the optimal value function to the solution of the GRLP. Our error bound contained two terms, one inherent to the ALP formulation and the other due to constraint reduction. We also made qualitative and quantitative observations about the nature of the error term that arose due to constraint reduction.
% Our analysis also revealed the fact that it is not always necessary to sample according to the stationary distribution of the optimal policy and, in fact, potentially several different constraint sampling/approximation strategies might work. 
In particular, we also theoretically justified linear function approximation of the constraints. We also discussed the results and provided a numerical example in the domain of controlled queues. To conclude, we observe that by providing a novel theoretical framework to study constraint approximation, this paper provides important results that add to the theory of ALP.
\end{comment}

### 2017-03-24 num.tex
exp, conc

%!TEX root =  autocontgrlp.tex
\section{Numerical Illustration}
In this section, we show via an example in the domain of controlled queues to show that the consequences of \Cref{conetheorm} (the \emph{cone} condition) derived based on the geometric intuition (please refer \cite{aaaipaper} for preliminary experiments in this domain).

\textbf{Queuing System Model:} The queuing model used here is similar to the one in Section~$5.2$ of \cite{ALP}. We consider a single queue with arrivals and departures. The state of the system is the queue length with the state space given by $\S=\{0,\ldots,S-1\}$, where $S-1$ is the buffer size of the queue. The action set $\A=\{1,\ldots,A\}$ is related to the service rates. We let $s_t$ denote the state at time $t$. The state at time $t+1$ when action $a_t \in \A $ is chosen is given by $s_{t+1}= s_{t}+1$ with probability $p$, $s_{t+1}= s_{t}-1$ with probability $q(a_t)$ and $s_{t+1}= s_t$, with probability $(1-p-q(a_t))$. For states $s_t=0$ and $s_t=n-1$, the system dynamics is given by $s_{t+1}= s_{t}+1$ with probability $p$ when $s_t=0$ and $s_{t+1}=s_t-1$ with probability $q(a_t)$ when $s_t=n-1$. The service rates satisfy $0<q(1)\leq \ldots\leq q(A)<1$ with $q(A)>p$ so as to ensure `stabilizability' of the queue. The reward associated with the action $a \in \A$ in state $s\in \S$ is given by $g_a(s)=-(s/n+q(a)^3)$.\par
\textbf{Parameter Settings:} We ran our experiments $S=1000$, $A=4$ with $q(1)=0.2$, $q(2)=0.4$, $q(3)=0.6$, $q(4)=0.8$, $p=0.4$ and $\alpha=1-\frac{1}{S}$.
The moderate size of $S=1000$ enabled us to compute the exact the exact value of $J^*$ (it was however expensive). We made use of polynomial features in $\Phi$ (i.e., $1,s,\ldots,s^{k-1}$) since they are known to work reasonably well for this domain \cite{ALP}, and chose $k=4$ (i.e., we used $1, s,s^2$ and $s^3$ as basis vectors).\\
\textbf{Experimental Methodology:} We compare two different sampling strategies $(i)$ based on the \emph{cone} conditions, and $(ii)$ based on constraint sampling. The two strategies are compared via  \emph{lookahead} policies, wherein, the action at state $s$ is obtained by computing the approximate value functions of the next states. The details are as follows:
$Case (i):$ Except for the corner states i.e., $s=1$ and $s=999$, each state $1<s<n-1$ has two next states namely $s'=s-1$ and $s'=s+1$. We formulate two separate LRALPs (or just one LRALP for $s=1$ and $s=n-1$) for next states. When formulating the LRALP for state $s'$, we let $c=e_{s'}$ and choose the constraint corresponding to state $s'$ to ensure the cone condition. We choose $5$ more constraints corresponding to states $1,200,400,600,800,999$ (uniformly spaced across the state space) and compute $\hat{J}_{e_s'}$. The lookahead policy is formulated as $u(s)=\underset{\arg\min}{a\in A}(g_a(s)+\underset{s'\in S}{\sum}p_a(s,s')\hat{J}_{e_s'}$.\\
$Case (ii):$ In a manner similar to $Case (i)$, here too we formulate two separate RLPs for next states. When formulating the RLP for state $s'$, we first let $c_{s'}(i)=(1-\alpha)(\alpha)^{|s'-i|}$ and normalize $c$ and sample $m=6$ constraints using $c$ and compute $\hat{J}_{c_s'}$. The lookahead policy is formulated as $u(s)=\underset{\arg\min}{a\in A}(g_a(s)+\underset{s'\in S}{\sum}p_a(s,s')\hat{J}_{c_s'}$.
\input{plt.tex}

>>>>

%!TEX root =  autocontgrlp.tex
\section{Numerical Illustration}
In this section, we show via an example in the domain of controlled queues the consequences of \cref{conetheorm},
which bounded the error when the constraints are chosen based on selecting a set of representative states
% (the \emph{cone} condition) derived based on the geometric intuition 
 (further preliminary experimental results have been reported in \cite{aaaipaper}).

\textbf{Model:} We run the experiments in the context of a queuing model similar to the one in Section~5.2 of \cite{ALP}. We consider a single queue with arrivals and departures. The state of the system is the queue length with the state space given by $\S=\{0,\ldots,S-1\}$, where $S-1$ is the buffer size of the queue. The action set $\A=\{1,\ldots,A\}$ is related to the service rates. We let $s_t$ denote the state at time $t$. The state at time $t+1$ when action $a_t \in \A$ is chosen is given by $s_{t+1}= s_{t}+1$ with probability $p$, $s_{t+1}= s_{t}-1$ with probability $q(a_t)$ and $s_{t+1}= s_t$, with probability $(1-p-q(a_t))$. For states $s_t=0$ and $s_t=n-1$, the system dynamics is given by $s_{t+1}= s_{t}+1$ with probability $p$ when $s_t=0$ and $s_{t+1}=s_t-1$ with probability $q(a_t)$ when $s_t=n-1$. The service rates satisfy $0<q(1)\leq \ldots\leq q(A)<1$ with $q(A)>p$ so as to ensure `stabilizability' of the queue. 
The reward associated with  action $a \in \A$ and state $s\in \S$ is given by $g_a(s)=-(s/n+q(a)^3)$.

\textbf{Parameter Settings:} We ran our experiments for $S=1000$, $A=4$ with $q(1)=0.2$, $q(2)=0.4$, $q(3)=0.6$, $q(4)=0.8$, $p=0.4$ and $\alpha=1-\frac{1}{S}$.
The moderate size of $S=1000$ enabled us to compute the exact value of $J^*$ (the most expensive part of the computation). 
Despite that the conic span condition cannot be met with them, as a way of testing the limits of the theory,
we made use of polynomial features in $\Phi$ (i.e., $1,s,\ldots,s^{k-1}$) 
since they are known to work reasonably well for this domain \cite{ALP}.
We chose $k=4$. % (i.e., we used $1, s,s^2$ and $s^3$ as basis vectors).

\textbf{Experimental Methodology:} We compare two different sampling strategies $(i)$ based on the \emph{cone} conditions, and $(ii)$ based on constraint sampling. The two strategies are compared via  \emph{lookahead} policies, wherein, the action at state $s$ is obtained by computing the approximate value functions of the next states. The details are as follows:
\emph{Case (i)}: Except for the corner states i.e., $s=0$ and $s=999$, each state $0<s<n-1$ has two next states namely $s'=s-1$ and $s'=s+1$. We formulate two separate LRALPs (or just one LRALP for $s=0$ and $s=n-1$) for next states. When formulating the LRALP for state $s'$, we let $c=e_{s'}$ and choose the constraint corresponding to state $s'$ to ensure the cone condition to be met for LRALP. We choose $5$ more constraints corresponding to states $1,200,400,600,800,999$ (uniformly spaced across the state space) and compute $\hat{J}_{e_s'}$. The lookahead policy is formulated as 
$u_{LRA}(s)=\argmin_{a\in A}(g_a(s)+\sum_{s'\in S}p_a(s,s')\hat{J}_{e_s'}(s')$.

\emph{Case (ii)}: In a manner similar to \emph{Case (i)}, here too we formulate two separate LRALPs for next states. When formulating the LRALP for state $s'$, we first let $c_{s'}(i)=(1-\alpha)(\alpha)^{|s'-i|}$ and normalize $c$ and sample $m=6$ constraints randomly using $c$ and compute $\hat{J}_{c_s'}$. 
The lookahead policy is formulated as 
$u_{CS}(s)=\argmin_{a\in A}(g_a(s)+\sum_{s'\in S}p_a(s,s')\hat{J}_{c_s'}(s')$.

The results are shown in the figures below:
\input{plt.tex}
The right-hand-side figure shows the policies computed, 
while the left-hand-side figure shows their value functions. 
Since constraint sampling (CS) produces randomized results,
we repeated the simulations 10 times. 
The results in all cases were quite similar except for one ``bad case'' for CS , hence we show the plot for a typical run.
As can be seen from the figure, choosing the constraints to (approximately) satisfy the constraint of the theoretical results
reliably produces better results: In fact, the value functions $J^*$ and $J_{LRA}$ are mostly on the top of each other.
While these results are only meant for illustration,
we expect that in larger domains it becomes even more important to select the constraints appropriately.
However, the study of this is left for future work.

### 2017-03-24 results.tex
in the middle
exp, conc

The purpose of this section is to present our main results.
\newcommand{\Jalpo}{J^*_{\alp}}
\newcommand{\Jlro}{J^*_{\lralpshort}}
\newcommand{\Jlr}{J_{\lralpshort}}
\newcommand{\rlr}{r_{\lralpshort}}
Let $\rlr$ be a solution to the LRALP given by \eqref{grlp}
and let $\Jlr = \Phi \rlr$. When multiple solutions exist, we can choose any of them.
For the result, we assume that the LRALP is not unbounded, and hence a solution exist. In fact, we will assume something much stronger. The discussion of why our assumptions are reasonable and how to ensure that they hold is postponed to after the presentation of our results.
Our main results bounds the error $\norm{J^* - \Jlr}_{1,c}$.

@@

Just like \citet{ALP}, we will also require that $\psi$ is a \emph{stochastic Lyapunov-function} for the MDP.  In particular, we require that the $\alpha$-discounted stability coefficient
\begin{align*}
\beta_\psi \doteq \alpha  \max_{a} \norm{P_a \psi}_{\infty,\psi}
\end{align*}
is strictly less than one.
This can be seen to imply that $H: (\Re^S,\norm{\cdot}_{\infty,\psi}) \to (\Re^{SA}, \norm{\cdot}_{\infty,\psi})$ is a contraction,
where for $J = (J_1^\top,\dots,J_A^\top)^\top \in \Re^{SA}$ we let $\norm{J}_{\infty,\psi} = \max_a\norm{J}_{\infty,\psi}$.
That $H$ is a contraction will play a crucial role in our results.
Note that the condition $\beta_\psi<1$ is closely related to the condition that for any policy $u$, 
$P_u \psi \le \psi$, which can be viewed as a stability condition on the MDP 
and which appeared in a slightly altered form in studying the stability of MDPs with infinite 
state spaces \citep[e.g.,][]{chemey99a}.
Note that one can always choose $\psi = \one$: $\beta_\one = \alpha<1$. 
With this, we are ready to state our main result:

@@

The next theorem shows that magnitudes of the coefficients used in the conic cover control the size of 
$\norm{\Jalpo-\Jlro}_{\infty,\psi}$. 
For the theorem we let $\Lambda\in \Re_+^{\S\times \S_0}$ be the matrix of conic coefficients: 
For any $s\in \S$, $\phi(s) = \sum_{s'\in \S_0} \Lambda(s,s') \phi(s')$. After the theorem we give constructions 
for creating conic covers.
\begin{theorem}
Assume that $W_1 = \dots = W_A$, $\{w_{1,a}\,dots,w_{m,a}\} = \{ e_s\,:\, s\in \S_0 \}$ 
and that $\phi(\S)$ lies in the conic span of $\phi(\S_0)$ with conic coefficients given by $\Lambda$.
Let $\epsilon = \inf_{r}\norm{ J^* -  \Phi r }_{\infty,\psi}$.
Then, 
\begin{align*}
\norm{\Jalpo-\Jlro}_{\infty,\psi}
\le 
\norm{\Jalpo-J^*}_{\infty,\psi} + (1+ \norm{\Lambda \psi}_{\psi,\infty}) \epsilon\,.
\end{align*}
\end{theorem}
\begin{proof}
Let $r^*$ be such that $\norm{J^* - \Phi r^*}_{\infty,\psi} = \epsilon$ (this exists by continuity) and let 
$\delta = J^* - \Phi r^*$.
Pick any $s\in \S$ and let $r_s = \argmin\{ r^\top \phi(s) \,:\, W^\top E \Phi r \ge W^\top E J^*,\, r\in \Re^k \}$ 
so that $\Jlro(s) = r_s^\top \phi(s)$.
Note that by assumption, for any $s'\in \S_0$, $\Jlro(s') = r_s^\top \phi(s') \ge J^*(s')$.
Now, notice that by definition, $\Jlro \le \Jalpo$ (the LP defining $\Jlro$ is the relaxation of the LP defining $\Jalpo$).
Hence,
\begin{align*}
0\le \Jalpo(s) - \Jlro(s) = \Jalpo(s) - J^*(s) + J^*(s) - \Jlro(s)
\end{align*}
and $\Jlro(s) = r_s^\top \phi(s) = r_s^\top \sum_{s'\in \S_0} \Lambda(s,s') \phi(s') = \sum_{s'\in \S_0} \Lambda(s,s') \Jlro(s')
\ge \sum_{s'\in \S_0} \Lambda(s,s') J^*(s')$.
Combining this with the previous inequality we get
\begin{align*}
0&\le 
 \frac{\Jalpo(s) - \Jlro(s)}{\psi(s)}\\
&\le
\frac{\Jalpo(s)-J^*(s)}{\psi(s)} + \frac{J^*(s) - \sum_{s'\in \S_0} \Lambda(s,s') J^*(s')}{\psi(s)}\,.
\end{align*}
Plugging in $J^*(s) = \phi(s)^\top r^*+\delta(s)$, using again that $\phi(s) = \sum_{s'\in \S_0} \Lambda(s,s') \phi(s')$,
and also using the triangle inequality after taking absolute values we get
\begin{align*}
\MoveEqLeft \frac{|J^*(s) - \sum_{s'\in \S_0} \Lambda(s,s') J^*(s')|}{\psi(s)} \\
& \le
  \frac{|\delta(s)|}{\psi(s)} + \frac{\sum_{s'\in \S_0} \Lambda(s,s') |\delta(s')|}{\psi(s)} \\
& \le
  \frac{|\delta(s)|}{\psi(s)} + \frac{1}{\psi(s)} \, \sum_{s'\in \S_0} \Lambda(s,s') \psi(s') \frac{|\delta(s')|}{\psi(s')} \\
& \le
  \frac{|\delta(s)|}{\psi(s)} + \norm{\delta}_{\infty,\psi} \frac{\sum_{s'\in \S_0} \Lambda(s,s') \psi(s')}{\psi(s)} \, .
\end{align*}
Combining this with the previous display and noting that $\norm{\delta}_{\infty,\psi} = \epsilon$ finishes the proof.
\if0
Putting things together, we get 
\begin{align*}
\norm{\Jalpo - \Jlro}_{\infty,\psi}
\le
\norm{\Jalpo(s)-J^*(s)}_{\infty,\psi}
+ 
\norm{\delta}_{\infty,\psi} (1+ \norm{\Lambda \psi}_{\infty,\psi})\,,
\end{align*}
finishing the proof.
\fi
\end{proof}

>>>>

The purpose of this section is to present our main results.
Let $\rlr$ be a solution to the LRALP given by \eqref{grlp}
and let $\Jlr = \Phi \rlr$. When multiple solutions exist, we can choose any of them.
For the result, we assume that the LRALP is not unbounded, and hence a solution exist. In fact, we will assume something much stronger. The discussion of why our assumptions are reasonable and how to ensure that they hold is postponed to after the presentation of our results.
Our main results bounds the error $\norm{J^* - \Jlr}_{1,c}$.

@@

Just like \citet{ALP}, we will also require that $\psi$ is a \emph{stochastic Lyapunov-function} for the MDP.  In particular, we require that the $\alpha$-discounted stability coefficient
\begin{align} \label{eq:betadef}
\beta_\psi \doteq \alpha  \max_{a} \norm{P_a \psi}_{\infty,\psi}
\end{align}
is strictly less than one.
This can be seen to imply that $H: (\Re^S,\norm{\cdot}_{\infty,\psi}) \to (\Re^{SA}, \norm{\cdot}_{\infty,\psi})$ is a contraction,
where for $J = (J_1^\top,\dots,J_A^\top)^\top \in \Re^{SA}$ we let $\norm{J}_{\infty,\psi} = \max_a\norm{J}_{\infty,\psi}$.
That $H$ is a contraction will play a crucial role in our results.
Note that the condition $\beta_\psi<1$ is closely related to the condition that for any policy $u$, 
$P_u \psi \le \psi$, which can be viewed as a stability condition on the MDP 
and which appeared in a slightly altered form in studying the stability of MDPs with infinite 
state spaces \citep[e.g.,][]{chemey99a}.
Note that one can always choose $\psi = \one$: $\beta_\one = \alpha<1$. 
With this, we are ready to state our main result:

@@

The next theorem shows that magnitudes of the coefficients used in the conic cover control the size of 
$\norm{\Jalpo-\Jlro}_{\infty,\psi}$. 
For the theorem we let $\Lambda\in \Re_+^{\S\times \S_0}$ be the matrix of conic coefficients: 
For any $s\in \S$, $\phi(s) = \sum_{s'\in \S_0} \Lambda(s,s') \phi(s')$. After the theorem we give constructions 
for creating conic covers.
\begin{theorem}
\label{conetheorm}
Assume that $W_1 = \dots = W_A$, $\{w_{1,a}\,dots,w_{m,a}\} = \{ e_s\,:\, s\in \S_0 \}$ 
and that $\phi(\S)$ lies in the conic span of $\phi(\S_0)$ with conic coefficients given by $\Lambda$.
Let $\epsilon = \inf_{r}\norm{ J^* -  \Phi r }_{\infty,\psi}$.
Then, 
\begin{align*}
\norm{\Jalpo-\Jlro}_{\infty,\psi}
\le 
\norm{\Jalpo-J^*}_{\infty,\psi} + (1+ \norm{\Lambda \psi}_{\psi,\infty}) \epsilon\,.
\end{align*}
\end{theorem}
\begin{proof}
Let $r^*$ be such that $\norm{J^* - \Phi r^*}_{\infty,\psi} = \epsilon$ (this exists by continuity) and let 
$\delta = J^* - \Phi r^*$.
Pick any $s\in \S$ and let $r_s = \argmin\{ r^\top \phi(s) \,:\, W^\top E \Phi r \ge W^\top E J^*,\, r\in \Re^k \}$ 
so that $\Jlro(s) = r_s^\top \phi(s)$.
Note that by assumption, for any $s'\in \S_0$, $\Jlro(s') = r_s^\top \phi(s') \ge J^*(s')$.
Now, notice that by definition, $\Jlro \le \Jalpo$ (the LP defining $\Jlro$ is the relaxation of the LP defining $\Jalpo$).
Hence,
\begin{align*}
0\le \Jalpo(s) - \Jlro(s) = \Jalpo(s) - J^*(s) + J^*(s) - \Jlro(s)
\end{align*}
and $\Jlro(s) = r_s^\top \phi(s) = r_s^\top \sum_{s'\in \S_0} \Lambda(s,s') \phi(s') = \sum_{s'\in \S_0} \Lambda(s,s') \Jlro(s')
\ge \sum_{s'\in \S_0} \Lambda(s,s') J^*(s')$.
Combining this with the previous inequality we get
\begin{align*}
0&\le 
 \frac{\Jalpo(s) - \Jlro(s)}{\psi(s)}\\
&\le
\frac{\Jalpo(s)-J^*(s)}{\psi(s)} + \frac{J^*(s) - \sum_{s'\in \S_0} \Lambda(s,s') J^*(s')}{\psi(s)}\,.
\end{align*}
Plugging in $J^*(s) = \phi(s)^\top r^*+\delta(s)$, using again that $\phi(s) = \sum_{s'\in \S_0} \Lambda(s,s') \phi(s')$,
and also using the triangle inequality after taking absolute values, we get
\begin{align*}
\MoveEqLeft \frac{|J^*(s) - \sum_{s'\in \S_0} \Lambda(s,s') J^*(s')|}{\psi(s)} \\
& \le
  \frac{|\delta(s)|}{\psi(s)} + \frac{\sum_{s'\in \S_0} \Lambda(s,s') |\delta(s')|}{\psi(s)} \\
%& \le
%  \frac{|\delta(s)|}{\psi(s)} + \frac{1}{\psi(s)} \, \sum_{s'\in \S_0} \Lambda(s,s') \psi(s') \frac{|\delta(s')|}{\psi(s')} \\
& \le
  \frac{|\delta(s)|}{\psi(s)} + \norm{\delta}_{\infty,\psi} \frac{\sum_{s'\in \S_0} \Lambda(s,s') \psi(s')}{\psi(s)} \, .
\end{align*}
Combining this with the previous display and noting that $\norm{\delta}_{\infty,\psi} = \epsilon$ finishes the proof.
\if0
Putting things together, we get 
\begin{align*}
\norm{\Jalpo - \Jlro}_{\infty,\psi}
\le
\norm{\Jalpo(s)-J^*(s)}_{\infty,\psi}
+ 
\norm{\delta}_{\infty,\psi} (1+ \norm{\Lambda \psi}_{\infty,\psi})\,,
\end{align*}
finishing the proof.
\fi
\end{proof}
